Time delay estimation in unresolved lensed quasars

### 2020-12-26 paper/aa-package/paper.tex
intro

As first pointed out in [\cite{Refsdal1964}], time-delay cosmography offers an alternative way of determining the Hubble parameter: the light rays coming from a distant source, e.g. a quasar, can be deflected by the gravitational field of an intervening massive object, e.g. a galaxy. If the field is strong enough, multiple images of the same source will be observed and, by tracking the light intensity of each image over time, a \emph{light curve} is obtained. Light curves associated with different images will exhibit a mutual time-delay ($\Delta T$), due to the different paths that the photons have travelled. As shown in [\cite{Refsdal1964}], this time-delay is related to the Hubble parameter as $H_0 \propto 1/\Delta T$. The major results obtained via time-delay cosmography come from the H0LiCOW collaboration [\cite{H0licow_XIII}], which has found $H_0 = 73.3_{-1.8}^{+1.7}$ km s$^{-1}$ Mpc$^{-1}$ from a sample of six lensed quasars monitored by the COSMOGRAIL  project [\cite{Cosmograil2020}], enhancing the tension up to $5.3 \sigma$. However, a more recent analysis of 40 strong gravitational lenses, from TDCOSMO+SLACS \cite{tdcosmo}, has found $H_0 = 67.4_{-3.2}^{+4.1}$ km s$^{-1}$ Mpc$^{-1}$, relaxing the Hubble tension and demonstrating the importance of understanding the lenses mass density profiles. This scenario motivates further studies aimed at improving the precision in the $H_0$ estimation. 

@@

In this paper we focus on the time delay estimation and its relative error. To date, a sample of about $220$ GLQs is available\footnote{https://research.ast.cam.ac.uk/lensedquasars/index.html}. However, only a very small subset, consisting of the brightests and with with well separated multiple images, has been used to measure $H_0$. The reason relies on the fact that it is easier and safer to extract information, and consequently to reduce the error on $H_0$, for such systems.

@@

In this respect, Fig. \ref{fig:qsosepmagboth} shows the distribution of the multiple images versus the maximum image separation for the known GLQs. Systems being part of the grey region, which represent a consistent fraction of the total sample, can be regarded as not optimal candidates, since they are less bright and/or have very small image separations. 

@@

	\caption[cm]{comments}

@@

In addition, [\cite{shu2020discovering}] have presented a new method to find GLQ systems from unresolved light curves. Such systems have therefore very small-separation between multiple images. This would make big telescopes the ideal instruments to perform lensed quasars monitoring, both in light of their high angular resolution and the geographical areas they are placed in, where the effects of atmospheric turbulence are less prominent. However, because of the time scales of the intrinsic variations of the source, such observation campaigns should last years [\cite{Cosmograil2020}]. Therefore, due to the amount of observational requests that big telescopes have to fulfil, they can not be employed for these purposes. On the other hand, small/medium sized telescopes ($1$-$2$m) [\cite{Borgeest1996}] can be used. Unfortunately, their reduced angular resolution is further worsened by their often less privileged geographical positions in terms of atmospheric seeing, up to 3 arcsec \cite{karttunen2016fundamental}. In addition, small telescopes cannot observe faint distant sources. 

@@

Here, we propose a novel approach based on Machine Learning (ML) algorithms to estimate the time-delay from non-resolved lensed quasar light curves. In particular, we train a one-dimensional Convolutional Neural Network (CNN) to map non-resolved double-lensed quasar images into the corresponding time-delay. Starting from open-source real data \cite{CosmograilData} of resolved quasar light curves, we combine them into a single time series to mimic the real situation in which the quasar images are not resolved. Then, we feed the so-obtained time series into a GP-based algorithm, which artificially generates new realistic non-resolved quasar light curves along with their corresponding time-delays. Preliminary experiments based on data of quasars HE0435 and HS2209 demonstrate the effectiveness of the proposed method.

%--------------------------------------------------------------------
\section{Light Curves Simulation}

>>>>

As first pointed out in [\cite{Refsdal1964}], time-delay cosmography offers an alternative way of determining the Hubble parameter: the light rays coming from a distant source, e.g. a quasar, can be deflected by the gravitational field of an intervening massive object, e.g. a galaxy. If the field is strong enough, multiple images of the same source are observed and, by tracking the light intensity of each image over time, a \emph{light curve} is obtained. Light curves associated with different images will exhibit a mutual time-delay ($\Delta T$), due to the different paths that the photons have travelled. As shown in [\cite{Refsdal1964}], this time-delay is related to the Hubble parameter as $H_0 \propto 1/\Delta T$. The major results obtained via time-delay cosmography come from the H0LiCOW collaboration [\cite{H0licow_XIII}], which has found $H_0 = 73.3_{-1.8}^{+1.7}$ km s$^{-1}$ Mpc$^{-1}$ from a sample of six lensed quasars monitored by the COSMOGRAIL  project [\cite{Cosmograil2020}], enhancing the tension up to $5.3 \sigma$. However, a more recent analysis of 40 strong gravitational lenses, from TDCOSMO+SLACS [\cite{tdcosmo}], has found $H_0 = 67.4_{-3.2}^{+4.1}$ km s$^{-1}$ Mpc$^{-1}$, relaxing the Hubble tension and demonstrating the importance of understanding the lenses mass density profiles. This scenario motivates further studies aimed at improving the precision in the $H_0$ estimation. 

@@

In this paper we focus on the time delay estimation and its relative error. 

@@

To date, a sample of about $220$ GLQs is available\footnote{https://research.ast.cam.ac.uk/lensedquasars/index.html}. However, only a very small subset, with well separated multiple images, has been used to measure $H_0$. The reason relies on the fact that it is easier and safer to extract information from such systems, and consequently reduce the error on $H_0$.
In this respect, Fig. \ref{fig:qsosepmagboth} (bottom) shows the magnitude of the multiple images versus the maximum image separation for the known GLQs. Systems being part of the grey region, which represent the 70$\%$ of the total sample (as shown in the up-right histogram), have a maximum image separation below 2 arcsec. 

@@

	\caption[cm]{Top-left: distribution of known GLQs as a function of the number of multiple images. Top-right: distribution of known GLQs as a function of the maximum angular separation. Bottom (left and right): Magnitude of the multiple images versus the maximum image separation. The grey region contains 70$\%$ of the total GLQ sample.}

@@

In addition, [\cite{shu2020discovering}] have presented a new method to find GLQs from unresolved light curves. Such systems have therefore even smaller separation between multiple images. 

@@

This would make big telescopes the ideal instruments to perform lensed quasars monitoring, both in light of their high angular resolution and the geographical areas they are placed in, where the effects of atmospheric turbulence are less prominent. However, because of the time scales of the intrinsic variations of the source, such observation campaigns should last years [\cite{Cosmograil2020}]. Therefore, due to the amount of observational requests that big telescopes have to fulfil, they can not be employed for these purposes. On the other hand, small/medium sized telescopes ($1$-$2$m) [\cite{Borgeest1996}] can be used. Unfortunately, their already reduced angular resolution is further worsened by their less privileged geographical positions, in terms of atmospheric seeing, which can reach 3 arcsec [\cite{karttunen2016fundamental}]. 
Therefore, the majority of GLQs already known, together with future discoveries, will mainly appear as a single image for small/medium-size telescopes. For this reason, here we propose a novel approach, based on Machine Learning (ML) algorithms, to estimate the time-delay from unresolved GLQ light curves. 
For simplicity, in this work we focus on double-lensed GLQs. This choice is further motivated by the fact that the majority of the already known systems are doubles, as shown in Fig. \ref{fig:qsosepmagboth} (top-left). However, [\cite{shu2020discovering}] expect to find a consistent quantity of quadruply-imaged QSOs in the future. Therefore, we plan to extend our work for systems with N>2 images.
The paper is structured as follows: Sec. \ref{sec:MC} describes the Monte Carlo (MC) simulations we performed....

train a one-dimensional Convolutional Neural Network (CNN) to map non-resolved double-lensed quasar images into the corresponding time-delay. Starting from open-source real data \cite{CosmograilData} of resolved quasar light curves, we combine them into a single time series to mimic the real situation in which the quasar images are not resolved. Then, we feed the so-obtained time series into a GP-based algorithm, which artificially generates new realistic non-resolved quasar light curves along with their corresponding time-delays. Preliminary experiments based on data of quasars HE0435 and HS2209 demonstrate the effectiveness of the proposed method.

%--------------------------------------------------------------------
\section{Light Curves Simulation} \label{sec:MC}
